\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\ or \\ Charting antigenic drift and evolutionary success of seasonal influenza lineages

### 2013-02-04 flux.tex
Implement revisions based on Colin's suggestions.

Through our analysis, we reveal the antigenic, as well as evolutionary, relationships among viruses in influenza A/H3N2, A/H1N1, B/Vic and B/Yam, quantifying both antigenic and evolutionary distances between strains (Figure~\ref{map}).
Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
We observe prominent antigenic clusters in A/H3N2 and A/H1N1, but less prominent, though still apparent, clustering in B/Vic and B/Yam.
Antigenic clusters show high genetic similarity, so that we observe very few mutation events leading to each cluster, rather than the repeated emergence of clusters.
This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Victoria/75 and Beijing/89) appearing to be evolutionary dead-ends.

@@

We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of $\beta=1.01$ antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.98--1.04.
However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
Most variation is contained within the first antigenic dimension, but dimension 2 occasionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing/89 and Beijing/92 clusters.

@@

Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift in the mean antigenic phenotype of the virus population is relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
For A/H3N2, the largest cluster transition we observe is the emergence of the Sydney/97 cluster, evident as the appearance in 1997 of a large gap in virus locations along antigenic dimension 1 (Figure~\ref{drift}).
However, in 1997 only three of the nine viruses sampled show the Sydney/97 phenotype, while the remaining six viruses exhibit the old Wuhan/95 phenotype.
In 1998, three of the four viruses sampled show the Sydney/97 phenotype, with the remaining virus still showing the Wuhan/95 phenotype.
In 1999, all sampled viruses exhibit the Sydney/97 phenotype.
Thus, we observe fairly equal rates of year-to-year drift in mean antigenic phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than all drift in mean antigenic phenotype attributable to a shift in 1997 (Figure~\ref{jumps}).
In addition to showing greater overall rates of antigenic drift, A/H3N2 shows greater variation in year-to-year drift than seen in A/H1N1 or influenza B (Figure~\ref{jumps}).
We find that the standard deviation in year-to-year drift is 0.88 antigenic units for A/H3N2, 0.51 units for A/H1N1, 0.54 units for B/Vic and 0.27 units for B/Yam.

>>>>

Through our analysis, we reveal the antigenic, as well as evolutionary, relationships among viruses in influenza A/H3N2, A/H1N1, B/Vic and B/Yam, quantifying both antigenic and evolutionary distances between strains (Figure~\ref{map}).
Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
We observe prominent antigenic clusters in A/H3N2 and A/H1N1, but less prominent, though still apparent, clustering in B/Vic and B/Yam.
Antigenic clusters show high genetic similarity, so that we observe very few mutation events leading to each cluster, rather than the repeated emergence of clusters.
This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Victoria/75 and Beijing/89) appearing to be evolutionary dead-ends.
Labeling of prominent antigenic clusters in figures \ref{map} and \ref{drift} is intended as a rough guide for orientation and not as exhaustive catalog of antigenic variation.

@@

We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of 1.01 antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.98--1.04 (Figure~\ref{drift}).
However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
Most variation is contained within the first antigenic dimension, but dimension 2 occasionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing/89 and Beijing/92 clusters.

@@

Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift in the mean antigenic phenotype of the virus population is relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
For A/H3N2, the largest cluster transition we observe is the emergence of the Sydney/97 cluster, evident as the appearance in 1997 of a large gap in virus locations along antigenic dimension 1 (Figure~\ref{drift}).
However, in 1997 only three of the nine viruses sampled show the Sydney/97 phenotype, while the remaining six viruses exhibit the old Wuhan/95 phenotype.
In 1998, three of the four viruses sampled show the Sydney/97 phenotype, with the remaining virus still showing the Wuhan/95 phenotype.
In 1999, all sampled viruses exhibit the Sydney/97 phenotype.
Thus, we observe fairly equal rates of year-to-year drift in mean antigenic phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than all drift in mean antigenic phenotype attributable to a shift in 1997 (Figure~\ref{jumps}).
This observation remains despite the fact that antigenic drift is accomplished, not by steady movement of the entire virus population, but by the successive emergence of drifted variants that come to take over the virus population.
In addition to showing greater overall rates of antigenic drift, A/H3N2 shows greater variation in year-to-year drift than seen in A/H1N1 or influenza B (Figure~\ref{jumps}).
We find that the standard deviation in year-to-year drift is 0.88 antigenic units for A/H3N2, 0.51 units for A/H1N1, 0.54 units for B/Vic and 0.27 units for B/Yam.
